Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-11 paper.tex
Add table for XE and move to appendix

% \begin{frame}
% \begin{block}{Multiple origin Pango X lineages}
% \begin{itemize}
% \item For the remaining 18, the story is more complicated.
% \item XJ, XU and XV each have 3 samples and 3 recombinants
% \item One recombinant shared among 5 lineages (XAA, XAB, XAG, XQ, XU)
% \item XE has 170 samples and 12 recombinants; some are very similar
% (Official: BA.1* \& BA.2, 10448--11287):
% \\
% \vspace{1em}
% \begin{tabular}{llll}
% \toprule
% Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
% \midrule
% BA.1.15 & BA.2  &9867  &10198\\
% BA.1.15 & BA.2  &9867  &10198\\
% BA.1.1  &BA.2  &10451  &11537\\
% BA.1.1.5 & BA.2 &10448 & 11537\\
% BA.1.15 & BA.2  &10448  &12880\\
% BA.1.17 & BA.2  &10448 & 11537\\
% BA.1  &BA.2.9  &10448 & 11537\\
% BA.1  &BA.2  &12764  &13195\\
% \bottomrule
% \end{tabular}
% % BA.3  &BA.1.17  &22674  &22674\\
% % BA.1.1.15 Recombinant 10448 11537
% % BA.1.1 Recombinant 10030 10447
% % Recombinant BA.2 17411 19955
% \end{itemize}
% \end{block}
% \end{frame}

\begin{figure}
\centering
\begin{tabular}{llll}
\toprule
Left parent & Right parent & \multicolumn{2}{l}{Breakpoint interval}\\
\midrule
BA.1.15  & BA.2   & 9,867  & 10,198 \\
BA.1.15  & BA.2   & 9,867  & 10,198 \\
BA.1.1   & BA.2   & 10,451 & 11,537 \\
BA.1.1.5 & BA.2   & 10,448 & 11,537 \\
BA.1.15  & BA.2   & 10,448 & 12,880 \\
BA.1.17  & BA.2   & 10,448 & 11,537 \\
BA.1     & BA.2.9 & 10,448 & 11,537 \\
BA.1     & BA.2   & 12,764 & 13,195 \\
\bottomrule
\end{tabular}
% BA.3  &BA.1.17  &22674  &22674\\
% BA.1.1.15 Recombinant 10448 11537
% BA.1.1 Recombinant 10030 10447
% Recombinant BA.2 17411 19955
\caption{\label{tab:multiple_origins_table}
Details of inferred
breakpoints for different recombination events inferred for XE samples (170
samples and 12 inferred recombination nodes; some are very similar (Official:
BA.1* \& BA.2, 10448–11287)}
\end{figure}

@@

As the results of \texttt{sc2ts} are sensitive to the existance of
time-travellers, an imporant aspect of future work is to find better ways
to identify them. One possibility is to use the LS HMM itself to flag
overly divergent sequences, and to exclude them from attachment to the ARG.
We might also estimate collection dates by adding these potential time travellers
back to the ARG, allowing an automated assessment of collection date discrepancies.

\subsection{Imputation of Pango lineage for non-sample nodes}
% FIXME the caption and the text are basically saying the same thing here. Can
% we streamline, and maybe give a bit more detail in the text about iteration,
% and the concrete implementation?
\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the iterative procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). The lineage for node N1 can be directly 
copied from its parent (L2). The lineage for node N2 must be inferred from 
that of the parent (L2) plus the lineage-defining mutation (red X) 
on the connecting edge. The lineage for node N3 can then be copied from 
that of node N2.}
\end{figure}
While the Pango lineages of sample nodes are imported directly from 
GISAID metadata, the lineage status of internal nodes inserted into 
the \texttt{sc2ts} ARG must be imputed. We do this using 
the list of lineage-defining mutations (based on 90\% consensus 
of the sequences analysed) from the COVID-CG website \citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.

@@

This method is fast, as the lineages for most of the internal nodes can be 
imputed by copying from the surrounding nodes (66\% of nodes for the Long ARG, 
80\% of nodes for the Wide ARG). The accuracy depends on
the quality of the list of lineage-defining mutations; we find that when
masking the lineages of sample nodes one-by-one and re-imputing these using
the remaining information results in 98\% (resp.~99\%) of lineages being
imputed correctly.

@@

>>>>

@@

As the results of \texttt{sc2ts} are sensitive to the existence of
time-travellers, an important aspect of future work is to find better ways
to identify them. One possibility is to use the LS HMM itself to flag
overly divergent sequences, and to exclude them from attachment to the ARG.
We might also estimate collection dates by adding these potential time travellers
back to the ARG, allowing an automated assessment of collection date discrepancies.

\subsection{Imputation of Pango lineage for non-sample nodes}

\begin{figure} \centering
\includegraphics[width=.7\textwidth]{figures/imputation.pdf}
\caption{\label{fig:imputation}A schematic of the iterative procedure to impute
Pango lineage for inserted, non-sample nodes. Here, three nodes (question
marks) have unknown Pango lineage (A). The lineage for node N1 can be directly
copied from its parent (L2). The lineage for node N2 must be inferred from
that of the parent (L2) plus the lineage-defining mutation (red X)
on the connecting edge. The lineage for node N3 can then be copied from
that of node N2.}
\end{figure}
While the Pango lineages of sample nodes are imported directly from
GISAID metadata, the lineage status of internal nodes inserted into
the \texttt{sc2ts} ARG must be imputed. We do this using
the list of lineage-defining mutations (based on 90\% consensus
of the sequences analysed) from the COVID-CG website \citep[][\url{https://covidcg.org/}; accessed on 4 November 2022]{Chen2021-zc}.

@@

This method is fast, as the lineages for most of the internal nodes can be
imputed by copying from the surrounding nodes (66\% of nodes for the long ARG,
80\% of nodes for the wide ARG), and significantly more efficient than extracting
the haplotypes for each internal node and using existing Pangolin assignment tools
\citep{OToole2021-assignment}. The
accuracy depends on the quality of the list of lineage-defining mutations,
as well as the source of lineage designation for the sample nodes: we obtain
slightly different results when using those recorded on GISAID (which uses
pangoLEARN), and those assigned by Nextclade. To gauge the accuracy of
imputation, we have used our method to re-impute the lineage designations of each
sample node using the surrounding information; this results in 98\% (resp.~99\%) of
nodes being assigned the same lineage as per the source metadata in the long
(resp.~wide) ARG.

@@

\begin{table}
\begin{tabular}{llll}
\toprule
 interval &                 causal & parent &                                descendant\_lineages \\
\midrule
 (9867, 10198) &             XE  & [BA.1, BA.2] &                                          \{'XE': 3\} \\
(10448, 11537) &             XH  & [BA.1.20, BA.2.9] &  \{'XAF': 34, 'XH': 11, 'B.1.1.529': 3, 'XE': 3,... \\
(10448, 11537) &             XE  & [BA.1.1.15, Unknown (R)] & \{'XE': 2\} \\
(10448, 11537) &             XE  & [BA.1.17.2, BA.2] &                                        \{'XE': 155\} \\
\bottomrule
\end{tabular}
\caption{\label{tab:multiple_origins_table}
Details of inferred
breakpoints for different recombination events inferred for XE samples
(Official: BA.1* \& BA.2, 10448–11287). TODO ELABORATE}
\end{table}
